When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-02-28 doc/paper/vpd_et_paper.tex
read through done, now just need to do checklist (figures, abstract, etc.)

The opposing effects of increased atmospheric demand and higher stomatal stress lead to two possible perspectives for how evapotranspiration (ET) responds to shifts in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand and fails to account for ecosystem response \citep{Swann_2016}. In reality, plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. This leads to a plant physiology perspective, in which an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure. This decrease in ET would reduce and potentially cancel, in the case of full stomatal closure, the effects of shifts to atmospheric demand. In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET response.

The ET response to changes in VPD alters water partitioning between the soil and atmosphere. If plant response reduces ET in response to atmospheric drying then soil moisture will be better conserved. This would seem a sensible evolutionary strategy to cope with aridity. If stomata were fully passive \citep [similar to soil pores, e.g. ][]{Or_2013}, increased atmospheric aridity would strongly reduce soil moisture \citep{Berg_2017}. In turn, this would further increase aridity as low soil moisture levels increase the Bowen ratio, and in turn increase temperature and atmospheric drying \citep[][]{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006, Salvucci_2013, Gentine_2016, Berg_2016}. This however would not seem to be a sensible strategy for plants from an evolutionary standpoint.

@@

The objective of the present manuscript is to evaluate the VPD dependence of ET, in non-extreme soil drought conditions. The goal of this paper is to use reasonable approximations as a tool to drive intuition for plant response to atmospheric drying. This intuition will aid interpretation of observations and full complexity models. In order to quantify plant response to perturbations in VPD, we apply a Penman-Monteith framework to derive the theoretical response of ET to VPD. The model is then validated and tested at multiple eddy-covariance stations spanning various climates and plant functional types.

@@

The purpose of this study is to examine ecosystem response to atmospheric drying focusing on the growing season. To accomplish this, we filter and quality control the data using a similar procedure as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data, according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the average Gross Primary Productivity (GPP) exceeds 10\% of the observed 95th percentile of GPP for a given site. GPP is calculated using the nighttime respiration partitioning method.
\item We remove days with rain and the day following to avoid issues with rain interception and sensor saturation at high relative humidity (\cite{MEDLYN_2011}).
\end{itemize}
Additionally, as in \citet{Lin_2018}, we restrict data to the daytime, which is identified when downwelling shortwave radiation is greater than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W m$^{-2}$. To reduce the chance of sensor saturation at high relative humidity, we remove all time steps for which VPD is less than .01 kPa, and to reduce errors at low windspeeds we remove all periods with wind magnitudes less than 0.5 m s$^{-1}$. Timesteps with negative observed GPP or ET are also removed, and we aggregate half hourly data to hourly averages to reduce noise \citep{Lin_2018}. After these quality control procedures, 335,939 upscaled hourly observations remain.

\subsection{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Penman_1948, Monteith_1965} estimates ET as a function of observable atmospheric variables and surface conductances:
\begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R_{net} + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with temperature, given by Clausius-Clapeyron ($\frac{d \; e_s}{d \; T}$), $R$ is the net radiation minus ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific heat of air at constant pressure, $\gamma$ is the psychometric constant, and $g_s$ is the stomatal conductance (Table \ref{definitions}).

@@

\begin{linenomath*}
  \begin{equation}
    uWUE = \frac{GPP \cdot \sqrt{VPD}}{ET}
    \label{uwue}
  \end{equation}
\end{linenomath*}
is relatively constant across time and moisture conditions within a plant functional type, in non-extreme drought conditions. If, following \citet{Lin_2015}, we approximate $g_0$ as $0$ (i.e. we neglect cuticular and epidermal losses - a reasonable assumption except in very dry conditions), we can use uWUE to remove the $A$ dependence from $g_s$ in a way that makes the PM equation analytically tractable:

\begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{uWUE \; ET}{c_a \; \sqrt{VPD}}
    \label{new_g_s}
  \end{equation}
\end{linenomath*}

@@

Given the eddy-covariance FLUXNET2015 data described in Section \ref{data}, every term in our new version of PM (Equation \ref{et}) is observed, except for the two parameters $g_1$, and uWUE. $g_1$ has been measured and reported for the PFTs considered here \citep{Lin_2015, Franks_2017}, and we fit uWUE by calculating its expectation, given the model and FLUXNET2015 data.

@@

Thus, with this uncertainty analysis we can evaluate departure from our theory in observations, as a departure of $\sigma$ from unity. The variability of $\sigma$ across sites and time provides a measure of uncertainty in our model, assumptions, as well as in the FLUXNET201 observations themselves. The variability of $\sigma$ then propagates through any uncertainty to our derivative of Equation \ref{et_sigma}:

@@

By construction, the variability in the $\sigma$ term (Equation \ref{sigma}) contains all model and observational uncertainties. For an observation that perfectly matches our model and constant uWUE assumption $\sigma$ will be one. Therefore, for our assumptions and framework to be reasonable $\sigma$ should be close to 1. An additional concern is that $\sigma$ may in fact be correlated with $VPD$, in which case the dependence would need to be accounted for when taking the derivative. Fortunately, there is a very weak dependence of $\sigma$ on VPD in their joint distrubution, and $\sigma$ is indeed close to unity i.e. $O(1)$ (Figure \ref{joint_vpd_sigma}). Given this weak dependence and the distribution of $\sigma$ we have confidence in our model framework and the data quality.

@@

\begin{equation}
  \label{sign}
  \frac{c_p}{R_{air}} - \frac{\gamma c_a }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right).
\end{equation}
All variables are positive, so the relative magnitude between the first term and the second term in the sign term (Equation \ref{sign}) will determine whether ET increases or decreases with increasing VPD. If the second term is larger then the plant control dominates and ET decreases with increasing VPD. However, if the first term is larger, then atmospheric demand dominates and ET increases with increasing VPD.

\subsection{Functional Form of the Sign Term}
\label{sign_term}
First, we explore the variables within the sign term to gain better intuition on the driver of either the increase or reduction of ET with VPD. CO$_2$ concentration ($c_a$) and the psychometric constant ($\gamma$) are relatively constant over the dataset considered here so that the variability is dominated by $\sigma$ and $VPD$. uWUE could vary with soil moisture but has been shown to be relatively constant \citep{Zhou_2015}. This then means that the sign term only depends on VPD for a given PFT and is approximately just a function of $VPD$. We can further determine a critical threshold separating an increase from a decrease in ET, i.e. the threshold $VPD_{crit}$ such as the derivative vanishes $\frac{\partial \; ET}{\partial \; VPD} = 0$:
\small
\begin{linenomath*}
  \begin{equation}
VPD_{crit} = \frac{R_{air}}{4 c_p} \left( \frac{\gamma c_a}{1.6\; R \;  uWUE} + \sqrt{\frac{\gamma c_a}{1.6\; R \;  uWUE}\left( \frac{\gamma c_a}{1.6\; R \;  uWUE} + 8 g_1 \frac{c_p}{R_{air}}\right)} - 4 g_1 \frac{c_p}{R_{air}} \right)
\label{vpd_min_et}
  \end{equation}
\end{linenomath*}
\normalsize
We note that $VPD_{crit}$ mostly depends on the PFT parameters uWUE and $g_1$ but only vary weakly with climate as most other parameters related to the environment are nearly constant. The value of $VPD_{crit}$ for each PFT is shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, ET will thus decrease with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} < 0$), and for values of $VPD$ greater than $VPD_{crit}$, ET will increase with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} > 0$). In other words, ecosystems regulate and mitigate evaporative losses up to the VPD limit, $VPD_{crit}$, above which atmospheric demand is just too high to be entirely compensated by stomatal and ecosystem regulation. We note however that even though ET increases again above the critical threshold, $VPD_{crit}$, ET is still much lower that potential evaporation as stomata are still strongly regulating vapor fluxes tot he atmosphere. Indeed, even in the absence of soil pore evaporation, stomata do not shut down entirely at very high VPD so that ET does not go to zero, because stomata are still slightly open to perform some photosynthesis \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. In addition, upward xylem transport is necessary to maintain phloem transport, as well as nutrient transport and thus carbon allocation \citep{De_2013, Nikinmaa_2013, Ryan_2014}.

@@

A tendency towards increasing ET response with increasing VPD can also be caused by a high slope parameter ($g_1$), characteristic of plants that at the leaf scale are more willing to trade water for stomata opening. Plants that are less conservative will be thus be more likely to increase ET with increasing VPD. Both the aforementioned effects (large uWUE, $g_1$) can amplify each other, and generally conspire to shift the sign term towards a positive value for a given PFT.

This effect of uWUE and $g1$ on the sign term is most apparent by comparing two extreme PFTs: water intensive crops (CRO) and water conservative closed shrub (CSH). CRO has high slope parameters and uWUE ($g_1 = 183.1$ Pa$^{1/2}$; $2.27$ $\mu$-mol [C] Pa$^{0.5}$ J$^{-1}$ [ET]) compared to CSH ($g1 = 148.6 \, Pa^{1/2}$, $uWUE=1.66$ $\mu$-mol [C] Pa$^{0.5}$ J$^{-1}$ [ET]). These differences in PFT parameters cause opposite ET responses to changes in VPD between CRO and CSH. ET theoretically always decreases with increasing VPD for the more water conservative CSH, while ET will frequently  increases with increasing VPD for the more water intensive CRO (Figure \ref{idealized_sign}). CRO evolved or were bred to prioritize GPP and yield and are thus not water conservative. They are very willing to trade water for photosynthesis and productivity, despite changes in VPD, while CSH are very unwilling to trade water for more photosynthesis.

@@

The control of the scaling term variability between PFTs by aerodynamic conductance is confirmed by the data (Figure \ref{scale_vary}). Differences between PFT are almost entirely due to differences in aerodynamic conductance, rather than differences in observed temperature ranges. The scaling term for the tree PFTs (DBF, EBF, ENF) is generally about double the scaling terms for the other PFTs which have lower surface roughness and generally smaller aerodynamic conductance.

@@

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
We now compare more sophisticated distributions of the observed response to our simplified theory (Section \ref{sign_term}). Figure \ref{test_sign} shows the observed distribution of the sign term, as compared to what the theory would predict. Our goal was to capture the leading order behavior of the ET dependence on VPD. Given the assumptions we have made, and the uncertainties of flux tower observations themselves, we expect a relatively large amount of noise when reproducing the derivatives of ET. However, the data largely reproduces our theoretical analysis.

@@

While CSH has a much different functional form of the sign term than DBF, CSH observations also match our theory to leading order. Again, the VPD dependence mostly determined from the slope parameter ($g_1$) closely matches the medians in the observation bins and the theoretical threshold is well reproduced. The VPD-independent, strongly negative response, which is determined by uWUE, is also captured. For CSH, there is extremely rare occurrence of observed positive ET response with VPD ($\approx 3$\%), even with uncertainty, so the sign of the observations almost always matches the sign of the theory, which states that ET response should always be negative.

As Table \ref{stats} suggested, the theory does not match as well for CRO as for DBF and CSH. At low VPD the VPD dependence (due to $g_1$) is similar to the leading order behavior of the observations, but the theory's curve is shifted towards more positive values relative to the observations, likely due to some bias in uWUE. Most strikingly, the observations show a tendency to move back towards negative values at high VPD, which given the functional form of VPD dependence, is behavior that is impossible for the theory to capture. This general positive bias of the theory explains the shift towards negative ET response observed in the bulk statistics in Table \ref{stats}. The divergence between the theory and the observations for CRO could have been expected. For practical reasons, our theory does not include important sources of plant type heterogeneity within and across sites for CRO. In agricultural landscape, the footprint of an eddy-covariance tend to not only observe a single species but many different ones, with potentially varying water use strategies. We also do not account for seasonal changes in LAI, variability in crop height and surface roughness, differences in C3 vs C4 photosynthesis and water strategies, or differences in irrigation practices between the sites and years. These deficiencies should largely explain the inability of our theory to exactly match the observations in croplands. For example, a superposition of sites with C3 plants (at low environmental VPD) and C4 plants (at high environmental VPD) would explain observed shift back towards negative ET response at high VPD when all sites are binned together, as in Figure \ref{test_sign}.

The theory for GRA suffers from similar limitations as for CRO. GRA observations are characterized by a consistent trend back towards negative ET response at higher VPD, which the functional form of our theory is incapable of accounting for. As compared to CRO, the divergence between the theory and the observations is even greater for GRA, but these differences are explainable using similar logic as we did for CRO. GRA also tends to exhibit substantial within-PFT variability, such as C3 vs C4 photosynthesis, as well as strong seasonal variability in height and LAI. As with CRO, the trend back towards negative ET response at higher VPD could be explained by a superposition of sites with C3 vs C4 sites. Unfortunately, because detailed site-level data of C3 vs C4 plant type is not available, we could not bin CRO and GRA further by irrigation and photosynthesis strategy. However, we hypothesize that the theory would validate against observations much better when photosynthesis pathways are accounted for.

The comparison between the theory and observations for ENF and EBF is unique from the rest of the PFTs. The VPD dependence of the theory, determined by the slope parameter ($g_1$), is not as well captured by observations. However, the theory still captures the general behavior and magnitude of the response, with some increase in error at low and/or high VPD.  We note that we used published constants for $g_1$ \citep[from ][]{Lin_2015, Franks_2017}, which determines the VPD dependence, and that deviations should be expected across sites. We could have fit our own constants so that the theory would best match the observations, but our goal was to use a priori results whenever possible in order to avoid over-fitting and to have a theoretical approach as independent to the observations as possible for further confirmation of the validity of the approach.

While the above discussion shows that our theory has some limitations when applied to some PFTs, especially for crops and grasslands, it does well for DBF and CSH PFTs, and captures the response at non-extreme VPD for ENF and EBF. It is useful to also compare our theory, the observed uncertainty in the ET response, and PET, as PET is often used as a proxy for ET in hydrometeorological studies and drought indices. Again, as in our discussion of the theoretical results, the PET sensitivity to VPD will be constant at about 3.5. In Figure \ref{test_sign} the ET response to shifts in dryness is always far smaller than 3.5 in the observations, even with all uncertainty included. So the PET response to atmospheric dryness is outside the range of all observable responses. This casts doubt on PET's usefulness for capturing the leading order behavior of water budget and dryness response at the land surface, such as used in several drought indices. However our theory (Equation \ref{et}) could be used with published uWUE and $g1$ data, as an alternative to PET to estimating land response and dryness, and would more accurately account for the response to atmospheric drying.

@@

Most of the results presented so far focus on the sign term, so now we turn to observations of ET response with the scaling term included (Figure \ref{data_scatter}). Until now, in the interest of developing leading order intuition for ET response, and to be conservative in our acknowledgment of model and observational error, we've considered $\sigma$ variability to be a measure of uncertainty. An alternative viewpoint is that $\sigma \cdot uWUE$ represents spatial and temporal variability of uWUE, which may be expected within bounds \citep[see Table \ref{pft}, also ][]{Zhou_2015}. This is a less conservative view; some of the $\sigma \cdot uWUE$ variability will be due to model and observational error so by viewing $\sigma \cdot uWUE$ as \textit{real} variability we run the risk of mistaking noise for signal. However, the advantage of this viewpoint is that, from \citet{Zhou_2014}, we have very high confidence that uWUE fit at the hourly timescale (as we do with $\sigma \cdot uWUE$) correctly captures the relationship between ET, GPP, and VPD, and that our form of PM introduces minimal error with its use of uWUE.

@@

As in Figure \ref{test_sign}, Figure \ref{data_scatter} matches our expectation based on simplified theory. The sign term is most strongly scaled by $g_a$, and in general the sign term shifts towards positive values as VPD observations increase. The willingness with which a given plant type evolved to use water dictates the occurrence of positive versus negative ET response. Water conservative plants are highly effective at mitigating the effects of atmospheric demand, and can store water for later use by reducing ET in response to increasing VPD.

>>>>

The opposing effects of increased atmospheric demand and higher stomatal stress lead to two possible perspectives for how evapotranspiration (ET) responds to shifts in VPD. The first, a hydrometeorological perspective, is that higher VPD increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies \citep[e.g.,][]{Heim_2002, Scheff_2015}, typically only quantifies changes in atmospheric demand and fails to account for ecosystem response \citep{Swann_2016}. In reality, plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to partially close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. This leads to a plant physiology perspective, in which an increase in VPD, particularly in well-watered soil conditions, may actually correspond to a decrease in ET because of stomatal closure.  In other words, the  question ``When does VPD drive or reduce ET?'' can be related to whether plant regulation or atmospheric demand dominates ET response.

The ET response to changes in VPD alters water partitioning between the soil and atmosphere. If plant response reduces ET with atmospheric drying then soil moisture will be better conserved. This would seem a sensible evolutionary strategy to cope with aridity. If stomata were fully passive \citep [similar to soil pores, e.g. ][]{Or_2013}, increased atmospheric aridity would strongly reduce soil moisture \citep{Berg_2017}. In turn, this would further increase aridity as low soil moisture levels increase the Bowen ratio, and cause increased temperature and atmospheric drying \citep[][]{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006, Salvucci_2013, Gentine_2016, Berg_2016}. This however would not seem to be a sensible strategy for plants from an evolutionary standpoint.

@@

The objective of the present manuscript is to evaluate the VPD dependence of ET, in non-extreme soil drought conditions. The goal of this paper is to use reasonable approximations as a tool to develop intuition for plant response to atmospheric drying. This intuition will aid interpretation of observations and full complexity models. In order to quantify plant response to perturbations in VPD, we apply a Penman-Monteith framework to derive the theoretical response of ET to VPD. The model is then validated and tested at multiple eddy-covariance stations spanning various climates and plant functional types.

@@

The purpose of this study is to examine ecosystem response to atmospheric drying, focusing on the growing season. To accomplish this, we filter and quality control the data using a similar procedure as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data, according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the average Gross Primary Productivity (GPP) exceeds 10\% of the observed 95th percentile of GPP for a given site. GPP is calculated using the nighttime respiration partitioning method.
\item We remove days with rain and the day following to avoid issues with rain interception and sensor saturation at high relative humidity (\cite{MEDLYN_2011}).
\end{itemize}
Additionally, as in \citet{Lin_2018}, we restrict data to the daytime, which is identified when downwelling shortwave radiation is greater than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W m$^{-2}$. To reduce the chance of sensor saturation at high relative humidity, we remove all time steps for which VPD is less than .01 kPa, and to reduce errors at low windspeeds we remove all periods with wind magnitudes less than 0.5 m s$^{-1}$. Timesteps with negative observed GPP or ET are also removed, and we aggregate half hourly data to hourly averages to reduce noise \citep{Lin_2018}. After these quality control procedures, 335,939 upscaled hourly observations remain.

\subsection{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Penman_1948, Monteith_1965} estimates ET as a function of observable atmospheric variables and surface conductances:
\begin{linenomath*}
  \begin{equation}
    \label{orig_pen}
    ET = \frac{\Delta R_{net} + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with temperature, given by Clausius-Clapeyron ($\frac{d \; e_s}{d \; T}$), $R_{net}$ is the net radiation minus ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific heat of air at constant pressure, $\gamma$ is the psychometric constant, and $g_s$ is the stomatal conductance (Table \ref{definitions}).

@@

\begin{linenomath*}
  \begin{equation}
    uWUE = \frac{GPP \cdot \sqrt{VPD}}{ET}
    \label{uwue}
  \end{equation}
\end{linenomath*}
is relatively constant across time and moisture conditions within a plant functional type. If, following \citet{Lin_2015}, we approximate $g_0$ as $0$ (i.e. we neglect cuticular and epidermal losses - a reasonable assumption except in very dry conditions), we can use uWUE to remove the $A$ dependence from $g_s$ in a way that makes the PM equation analytically tractable:

\begin{linenomath*}
  \begin{equation}
    g_s = \frac{R \, T}{P} 1.6 \left(1 + \frac{g_1}{\sqrt{VPD}}\right) \frac{uWUE \; ET}{c_a \; \sqrt{VPD}}.
    \label{new_g_s}
  \end{equation}
\end{linenomath*}

@@

Given eddy-covariance FLUXNET2015 data (Section \ref{data}), every term in our new version of PM (Equation \ref{et}) is observed, except for the two parameters $g_1$, and uWUE. $g_1$ has been measured and reported for the PFTs considered here \citep{Lin_2015, Franks_2017}, and we fit uWUE by calculating its expectation, given the model and FLUXNET2015 data.

@@

Thus, with this uncertainty analysis we can evaluate departure from our theory in observations, as a departure of $\sigma$ from unity. The variability of $\sigma$ across sites and time provides a measure of uncertainty in our model, assumptions, as well as in the FLUXNET2015 observations themselves. The variability of $\sigma$ then propagates through any uncertainty to our derivative of Equation \ref{et_sigma}:

@@

By construction, the variability in the $\sigma$ term (Equation \ref{sigma}) contains all model and observational uncertainties. For an observation that perfectly matches our model and constant uWUE assumption, $\sigma$ will be one. Therefore, for our assumptions and framework to be reasonable $\sigma$ should be close to 1. An additional concern is that $\sigma$ may in fact be correlated with $VPD$, in which case the dependence would need to be accounted for when taking the derivative. Fortunately, there is a very weak dependence of $\sigma$ on VPD in their joint distrubution, and $\sigma$ is indeed close to unity i.e. $O(1)$ (Figure \ref{joint_vpd_sigma}). Given this weak dependence and the distribution of $\sigma$ we have confidence in our model framework and the data quality.

@@

\begin{equation}
  \label{sign}
  \frac{c_p}{R_{air}} - \frac{\gamma c_a }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right).
\end{equation}
All variables are positive, so the relative magnitude between the first term and the second term in the sign term (Equation \ref{sign}) will determine whether ET increases or decreases with increasing VPD. If the second term is larger then plant control dominates and ET decreases with increasing VPD. However, if the first term is larger, then atmospheric demand dominates and ET increases with increasing VPD.

\subsection{Functional Form of the Sign Term}
\label{sign_term}
First, we explore the variables within the sign term to gain better intuition on the driver of either the increase or reduction of ET with VPD. CO$_2$ concentration ($c_a$) and the psychometric constant ($\gamma$) are relatively constant over the dataset considered here so that the variability is dominated by $\sigma$ and $VPD$. uWUE could vary with soil moisture but has been shown to be relatively constant \citep{Zhou_2015}. This then means that the sign term only depends on VPD for a given PFT and is approximately just a function of $VPD$. We can further determine a critical threshold separating an increase from a decrease in ET, i.e. the threshold $VPD_{crit}$ such that the derivative vanishes $\frac{\partial \; ET}{\partial \; VPD} = 0$:
\small
\begin{linenomath*}
  \begin{equation}
VPD_{crit} = \frac{R_{air}}{4 c_p} \left( \frac{\gamma c_a}{1.6\; R \;  uWUE} + \sqrt{\frac{\gamma c_a}{1.6\; R \;  uWUE}\left( \frac{\gamma c_a}{1.6\; R \;  uWUE} + 8 g_1 \frac{c_p}{R_{air}}\right)} - 4 g_1 \frac{c_p}{R_{air}} \right),
\label{vpd_min_et}
  \end{equation}
\end{linenomath*}
\normalsize
noting that $VPD_{crit}$ mostly depends on the PFT parameters uWUE and $g_1$, and only varies weakly with climate as most other parameters related to the environment are nearly constant. The calculated value of $VPD_{crit}$ for each PFT is shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, ET will thus decrease with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} < 0$), and for values of $VPD$ greater than $VPD_{crit}$, ET will increase with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} > 0$). In other words, ecosystems regulate and mitigate evaporative losses up to the VPD limit, $VPD_{crit}$, above which atmospheric demand is just too high to be entirely compensated by stomatal and ecosystem regulation. We note however that even though ET increases again above the critical threshold, $VPD_{crit}$, ET is still much lower that potential evaporation as stomata are still strongly regulating vapor fluxes to the atmosphere. Indeed, even in the absence of soil pore evaporation, stomata do not shut down entirely at very high VPD so that ET does not go to zero, because stomata are still slightly open to perform some photosynthesis \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. In addition, upward xylem transport is necessary to maintain phloem transport, as well as nutrient transport and thus carbon allocation \citep{De_2013, Nikinmaa_2013, Ryan_2014}.

@@

A tendency towards increasing ET response with increasing VPD can also be caused by a high slope parameter ($g_1$), characteristic of plants that at the leaf scale are more willing to trade water for access to atmospheric CO$_2$. Plants that are less conservative will be thus be more likely to increase ET with increasing VPD. Both the aforementioned effects (large uWUE, $g_1$) can amplify each other, and generally conspire to shift the sign term towards a positive value for a given PFT.

This effect of uWUE and $g1$ on the sign term is most apparent by comparing two extreme PFTs: water intensive crops (CRO) and water conservative closed shrub (CSH). CRO has high slope parameters and uWUE ($g_1 = 183.1$ Pa$^{1/2}$; $2.27$ $\mu$-mol [C] Pa$^{0.5}$ J$^{-1}$ [ET]) compared to CSH ($g1 = 148.6 \, Pa^{1/2}$, $uWUE=1.66$ $\mu$-mol [C] Pa$^{0.5}$ J$^{-1}$ [ET]). These differences in PFT parameters cause opposite ET responses to changes in VPD between CRO and CSH. ET theoretically always decreases with increasing VPD for the more water conservative CSH, while ET frequently  increases with increasing VPD for the more water intensive CRO (Figure \ref{idealized_sign}). CRO evolved or were bred to prioritize GPP and yield and are thus not water conservative. They are very willing to trade water for photosynthesis and productivity, despite changes in VPD, while CSH are very unwilling to trade water for more photosynthesis.

@@

The control of the scaling term variability between PFTs by aerodynamic conductance is confirmed by data (Figure \ref{scale_vary}). Differences between PFT are almost entirely due to differences in aerodynamic conductance, rather than differences in observed temperature ranges. The scaling term for the tree PFTs (DBF, EBF, ENF) is generally about double the scaling terms for the other PFTs which have lower surface roughness and generally smaller aerodynamic conductance.

@@

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
We now compare more sophisticated distributions of the observed response to our simplified theory (Section \ref{sign_term}). The observed distribution of the sign term, as compared to what the theory would predict, is provided in Figure \ref{test_sign}. Our goal was to capture the leading order behavior of the ET dependence on VPD. Given the assumptions we have made, and the uncertainties of flux tower observations themselves, we expect a relatively large amount of noise when reproducing the derivatives of ET. However, the data largely reproduces our theoretical analysis.

@@

While CSH has a much different functional form of the sign term than DBF, CSH observations also match our theory to leading order. Again, the VPD dependence mostly determined from the slope parameter ($g_1$) closely matches the medians in the observation bins and the theoretical threshold is well reproduced (albeit off-scale). The VPD-independent, strongly negative response, which is determined by uWUE, is also captured. For CSH, there is extremely rare occurrence of observed positive ET response with VPD ($\approx 3$\%), even with uncertainty, so the sign of the observations almost always matches the sign of the theory, which states that ET response should always be negative.

As Table \ref{stats} suggested, the theory does not match as well for CRO as for DBF and CSH. At low VPD the VPD dependence (due to $g_1$) is similar to the leading order behavior of the observations, but the theory's curve is shifted towards more positive values relative to the observations, likely due to some bias in uWUE. Most strikingly, the observations show a tendency to move back towards negative values at high VPD, which given the functional form of VPD dependence, is behavior that is impossible for the theory to capture. This general positive bias of the theory explains the shift towards negative ET response observed in the bulk statistics in Table \ref{stats}. The divergence between the theory and the observations for CRO could have been expected. For practical reasons, our theory does not include important sources of plant type heterogeneity within and across sites for CRO. In agricultural landscape, the footprint of an eddy-covariance towers tend to not only observe a single species but many different ones, with potentially varying water use strategies. We also do not account for seasonal changes in LAI, variability in crop height and surface roughness, differences in C3 vs C4 photosynthesis and water strategies, or differences in irrigation practices between the sites and years. These deficiencies should largely explain the inability of our theory to exactly match the observations in croplands. For example, a superposition of sites with C3 plants (at low environmental VPD) and C4 plants (at high environmental VPD) would explain observed shift back towards negative ET response at high VPD when all sites are binned together, as in Figure \ref{test_sign}.

The theory for GRA suffers from similar limitations as for CRO. GRA observations are characterized by a consistent trend back towards negative ET response at higher VPD, which the functional form of our theory is incapable of accounting for. As compared to CRO, the divergence between the theory and the observations is even greater for GRA, but these differences are explainable using similar logic as we did for CRO. GRA also tends to exhibit substantial within-PFT variability, such as C3 vs C4 photosynthesis, as well as strong seasonal variability in height and LAI. As with CRO, the trend back towards negative ET response at higher VPD could be explained by a superposition of sites with C3 vs C4 plants. Unfortunately, because detailed site-level data of C3 vs C4 plant type is not available, we could not bin CRO and GRA further by irrigation and photosynthesis strategy. However, we hypothesize that the theory would validate against observations much better when photosynthesis pathways are accounted for.

The comparison between the theory and observations for ENF and EBF is unique from the rest of the PFTs. The VPD dependence of the theory, determined by the slope parameter ($g_1$), is not as well captured by observations. However, the theory still captures the general behavior and magnitude of the response, with some increase in error at low and/or high VPD.  We note that we used published constants for $g_1$ \citep[from ][]{Lin_2015, Franks_2017}, which determines the VPD dependence, and that deviations should be expected across sites \citep{Lin_2015}. We could have fit our own constants so that the theory would best match the observations, but our goal was to use a priori results whenever possible in order to avoid over-fitting and to have a theoretical approach as independent to the observations as possible for further confirmation of the validity of the approach.

While the above discussion shows that our theory has some limitations when applied to some PFTs, especially for crops and grasslands, it does well for DBF and CSH PFTs, and captures the response at non-extreme VPD for ENF and EBF. It is useful to also compare our theory, the observed uncertainty in the ET response, and PET, as PET is often used as a proxy for ET in hydrometeorological studies and drought indices. Again, as in our discussion of the theoretical results, the PET sensitivity to VPD will be constant at about 3.5. In Figure \ref{test_sign} the ET response to shifts in dryness is always far smaller than 3.5 in the observations, even with all uncertainty included. So the PET response to atmospheric dryness is outside the range of all observable responses. This casts doubt on PET's usefulness for capturing the leading order behavior of water budget and dryness response at the land surface, as is done in several drought indices. However our theory (Equation \ref{et}) could be used with published uWUE and $g1$ data, as an alternative to PET to estimating land response and dryness, and would more accurately account for the response to atmospheric drying.

@@

Most of the results presented so far focus on the sign term, so now we turn to observations of ET response with the scaling term included (Figure \ref{data_scatter}). Until now, in the interest of developing leading order intuition for ET response, and to be conservative in our acknowledgment of model and observational error, we've considered $\sigma$ variability to be a measure of uncertainty. An alternative viewpoint is that $\sigma \cdot uWUE$ represents spatial and temporal variability of uWUE, which may be expected within bounds \citep[see Table \ref{pft}, also ][]{Zhou_2015}. This is a less conservative view; some of the $\sigma \cdot uWUE$ variability will be due to model and observational error, so by viewing $\sigma \cdot uWUE$ as \textit{real} variability we run the risk of mistaking noise for signal. However, the advantage of this viewpoint is that, from \citet{Zhou_2014}, we have very high confidence that uWUE fit at the hourly timescale (as we do with $\sigma \cdot uWUE$) correctly captures the relationship between ET, GPP, and VPD, and that our form of PM introduces minimal error with its use of uWUE.

@@

As in Figure \ref{test_sign}, Figure \ref{data_scatter} matches our expectation based on simplified theory. The sign term is most strongly scaled by $g_a$, and in general the sign term shifts towards positive values as VPD increases. The willingness with which a given plant type evolved to use water dictates the occurrence of positive versus negative ET response. Water conservative plants are highly effective at mitigating the effects of atmospheric demand, and can store water for later use by reducing ET in response to increasing VPD.
